---
title: "Essential R to learn for statistics and data science in 2024"
author: "Joe Paris"
output: html_notebook
---

Essential R concepts and tidyverse programming paradigm.

## Basic Data Types

### Numeric data

Integers or floating-point numbers

### Character data

Strings

### Logical data

Booleans

### Factor data

Looks like a string but underneath it's an integer.

Conceptually similar to an enum.

Used to represent categorical data.

Map labels to "levels".

|         |         |             |             |
|---------|---------|-------------|-------------|
| Labels: | Placebo | Treatment 1 | Treatment 2 |
| Levels  | 1       | 2           | 3           |

## Data Structures

### Vectors

Ordered set of items with the same type.

``` r
c(1, 2, 3, 4, 5)
```

`c()` is the combine or concatenate function,

Elements are accessed via a 1-based indices.

```{r}
vec = c(1, 2, 3, 4, 5)
vec[1]
```

### Matrices

2-dimensional *numerical* data.

*Column major order*

```{r}
mat = matrix(1:9, nrow = 3, ncol = 3)
mat
```

### Arrays

Higher-dimensional *numerical* data structure.

Column, row, z

``` r
array()
```

### Lists

Lists allow for storing different data types in the same structure.

Elements can be accessed via indices or key/value pairs.

```{r}
print("indices")

list1 = list(37, 3.14, "Bob")

list1[1]
list1[2]
list1[3]
list1[1+1]

print("key/value pairs")

list2 = list(
  "age" = 37,
  "pi" = 3.14,
  "name" = "Bob"
)

list2["name"]
list2[3]
list2$name
```

### DataFrames

DataFrames have the same shape as matrices but each column can be of a different data type.

```{r}
df = data.frame(
  names = c("Frodo", "Sam", "Pippin", "Merry"),
  values = c(37, 42, 59, 8)
)

df
```

### Tibbles

The tidyverse uses [Tibbles](https://tibble.tidyverse.org/) instead of DataFrames.

Tibbles are enhanced DataFrames.

> A tibble, or `tbl_df`, is a modern reimagining of the data.frame, keeping what time has proven to be effective, and throwing out what is not. Tibbles are data.frames that are lazy and surly: they do less (i.e. they don’t change variable names or types, and don’t do partial matching) and complain more (e.g. when a variable does not exist). This forces you to confront problems earlier, typically leading to cleaner, more expressive code. Tibbles also have an enhanced `print()` method which makes them easier to use with large datasets containing complex objects.
>
> If you are new to tibbles, the best place to start is the [tibbles chapter](https://r4ds.had.co.nz/tibbles.html) in R for data science.

```{r}
pacman::p_load("tidyverse")

df = tibble(
  names = c("Frodo", "Sam", "Pippin", "Merry"),
  values = c(37, 42, 59, 8)
)

df
```

## Iteration

* `for` loops: useful when you know how many times you need to iterate
* `while` loops: useful when you don's or can't know how many times you will need to iterate (such as when performing an optimization)

`while` loops execute while some condition holds true.

```{r}
counter <- 1

while (counter <= 10) {
  print(counter)
  counter <- counter + 1
}
```

`for` loops execute the same code over every element of a given set.

```{r}
for (number in 1:10) {
  print(number)
}
```
You will often store intermediate results in another data structure.

```{r}
cumulative_sums <- c()
running_total <- 0
for (number in 1:10) {
  running_total = running_total + number
  cumulative_sums <- c(cumulative_sums, running_total)
}
print(cumulative_sums)
```

`for` and `while` come in "base R." There are packages that add additional looping structures, such as [foreach](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html).
